Tools for identifying implausible counts

### 2019-04-03 Rmd/count_check_MS.Rmd
added chironomids

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = TRUE, warning = FALSE, message = FALSE)
options(digits = 1)

@@

loadd(testate_counts)
loadd(testate_summary)
```

@@

A common assertion in papers reporting microfossil assemblage data is that a minimum of _N_ microfossils were counted, where _N_ is often 50 for chironomids and several hundred for pollen and diatoms. This is important because larger count sums are associated with smaller uncertainties, both is in relative abundance of taxa and derived statistics such are transfer function reconstructions [@Heiri2001]. However, given the time-consuming nature of microfossil counting, especially when preservation is poor or concentrations are low, there is an incentive to misreport the minimum count sum.

@@

The key insight that allows inference about the count sum is that assemblage data are expected to follow the typical features of a rank abundance curve. In particular, because there are many rare taxa in most communities [@Darwin1859], most community or assemblage samples will include taxa represented by a single individual [@Coddington2009], hereafter singletons, unless the sampling effort is high relative to the species richness. In the occasional assemblages without singletons, the greatest common divider should usually be one, especially if species richness is high.

@@

Data sets were downloaded from the Palaeodata Center, Neotoma, Pangaea and other sources. A range of microfossil data were sought to allow for difference in typical count sum and species richness. This paper does not attempt to be an exhaustive survey of all the data available.

@@

Dataset 1 consists of `r #nrow(d1)` chironomid samples from REF [-@REF]. As the count sums are given for each sample, the data have been back-transformed to count data. Each sample `r #stopifnot(FALSE)` has at least one singleton (whole or half chironomid head capsule). The mean number of singletons is `r "##MEAN"`. Assuming that the number of singletons comes from an approximately Poisson distribution (Fig \@ref(fig:singleton-hist)), the expected proportion of samples without singletons is `r #estnos`.

Dataset 2 is the `r n_distinct(testate_counts$sampleID)` sample North American testate amoeba training set with available counts collated by Amesbury et al. [-@AMESBURY2018]. The count sums range between `r testate_summary$count_summary$sum %>% range() %>% paste(collapse = " and ")` (median = `r testate_summary$count_summary$sum %>% median()`) tests, and species richness ranges between `r testate_summary$count_summary$n %>% range() %>% paste(collapse = " and ")` (median = `r testate_summary$count_summary$n %>% median()`) species. `r testate_summary$prop_singletons * 100`% of the assemblages have at least one singleton (median `r testate_summary$median_singletons`). In the samples that lack singletons, `r testate_summary$gcd1 * 100`% have a greatest common divider of one.

#predictors of singletons
#description of samples with gcd > 1

@@

Dataset 1. reconstructing reported percent. Problem with halves.

Dataset 4. Larocque-Tobler et al (2015) originally archived counts

Dataset 5 diatoms Siberia

@@

relationship to GRIM

@@

Low diversity

@@

Some taxonomic groups have microfossils that often come in groups of attached individuals. For example, for diatoms valves are counted, but these are commonly paired or in long chains.

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = TRUE, warning = TRUE, message = TRUE)

#set ggplot theme
theme_set(theme_bw(base_size = 12))
options(digits = 1)

@@

loadd(bird_singleton_summary)
loadd(testate_counts, testate_summary)
loadd(last_chance_meta)
```

@@

A common assertion in papers reporting microfossil assemblage data is that a minimum of _N_ microfossils were counted, where _N_ is often 50 for chironomids and several hundred for pollen and diatoms. This is important because larger count sums are associated with smaller uncertainties, both is in relative abundance of taxa and derived statistics such are transfer function reconstructions [@Heiri2001]. However, given the time-consuming nature of microfossil counting, especially when preservation is poor or concentrations are low, there is an incentive to misreport the minimum count sum. There is a documented case of mis-reported count sums: Larocque-Tobler et al [@Larocque-Tobler2015] reported that their chironomid count sums were at least 50 head capsules, a subsequent corrigendum  [@Larocque-Tobler2016] acknowledged that count sums were actually as low as nineteen. Telford [-@Telford2019] reports some other cases where count sums may be much lower than reported.

@@

The key insight that allows inference about the count sum is that assemblage data are expected to follow the typical features of a rank abundance curve. In particular, because there are many rare taxa in most communities [@Darwin1859], most community or assemblage samples will include taxa represented by a single individual [@Coddington2009], hereafter singletons, unless the sampling effort is high relative to the species richness. In the occasional assemblages without singletons, the greatest common divisor should usually be one, especially if species richness is high.

@@

Data sets were downloaded from the Palaeodata Center, Neotoma, Pangaea and other sources. A range of ecological and palaeoecological data were sought to allow for difference in typical count sum and species richness. This paper does not attempt to be an exhaustive survey of all the data available.

@@

The first dataset consists of over `r format(floor(bird_singleton_summary$n_routes[1]/1000) * 1000, big.mark = ",")` bird counts from the North American breeding bird survey [@NAmerican_Breeding_Bird2018]. The vast majority of counts have singletons species (Table \@ref(tab:bird-table))). To explore the effect of taxonomic richness on the prevalence of singletons, I aggregate the birds counts to progressively lower taxonomic resolution. As richness declines, the proportion of counts lacking singletons increases (Table \@ref(tab:bird-table))), reaching a moderate proportion at the order level, where most of the counts lacking singletons have fewer than five taxa.

The vast majority of counts without singletons have a greatest common divisor of one, except when taxonomic richness is low.

```{r bird-table}
bird_singleton_summary %>%
  select(`Taxonomic level` = taxonomic_level, `Percent without singletons` = pc_singletons, `Percent GCD > 1` = GCD, `Mean richness` = mean_richness) %>% 
  knitr::kable(digits = c(0, 4, 4, 3), caption = "Percent of bird counts without singletons and of greatest common divisor (GCD) greater than one of counts without singletons, at different taxonomic levels.")
```

To test the sensitivity of the prevalence of counts without singletons, I re-sample the counts with 400 or more observations to smaller count sums using the multivariate hypergeometric distribution in the extraDistr package version `r packageVersion("extraDistr")` [@extraDistr]. At the order level, the proportion of counts without singletons declines steeply until it reaches a minimum at about 200, thereafter is rises slowly (Fig. \@ref(fig:bird-recount)). The prevalence of counts without singleton is high for small counts because it is possible that only the common taxa are counted, with larger counts, the chance of counting a single individual of a rare taxon increases. With even larger counts the prevalence of counts without singletons increases due to saturation as singletons become doubletons and few new taxa are found.

```{r bird-recount, fig.cap="Effect of count sum on the proportion of samples without singletons. Results are the mean of ten trials, error bars are two standard errors.", fig.width = 90/25.4, fig.height = 90/25.4}
readd(bird_recount_singleton_plot)
```

The second dataset is the `r n_distinct(testate_counts$sampleID)` sample North American testate amoeba training set with available counts collated by Amesbury et al. [-@AMESBURY2018]. The count sums range between `r testate_summary$count_summary$sum %>% range() %>% paste(collapse = " and ")` (median = `r testate_summary$count_summary$sum %>% median()`) tests, and species richness ranges between `r testate_summary$count_summary$n %>% range() %>% paste(collapse = " and ")` (median = `r testate_summary$count_summary$n %>% median()`) species. `r testate_summary$prop_singletons * 100`% of the assemblages have at least one singleton (median `r testate_summary$median_singletons`). In the assemblages that lack singletons, `r testate_summary$gcd1 * 100`% have a greatest common divider of one.

<!-- description of samples with gcd > 1 -->

## Estimating the percent sum

The third dataset consists of `r nrow(last_chance_meta)` chironomid head capsule counts from Last Chance Lake [@Axford2017]. It includes both the count sums and the percentage of each taxa, allowing the count sums estimated from the rarest taxon to be checked. The percent are given to two decimal places: to account for rounding errors, I use the smallest and largest values consistent with the reported percentage and show this range as a error bar. All the estimated count sums are, within error, either identical to the reported count sum or twice as much (Fig. \@ref(fig:last-chance-plot)). The factor of two difference is because in some counts the rarest taxon is represented by a half head capsule. Reporting half microfossils is common for several microfossil groups, including chironomids and pollen (especially for bisaccate conifers), occasionally other fractions are reported. Fractional counts make the estimated count too high, so do no risk incorretly flagging counts as being too small.

```{r last-chance-plot, fig.cap = "Estimated and reported chironomid count sums from Axford et al. [-@Axford2017]. Lines show the 1:1 and 2:1 relationships.", fig.width = 90/25.4, fig.height = 90/25.4}
readd(last_chance_plot)
```

## Unexpected count data

@@

## Unexpected percent data
Dataset 5 diatoms Siberia

@@

Chao index
relationship to GRIM

@@

different count sums (eg trees & herbs vs trees herbs and pteridiophytes)

Low diversity 
Truncated diversity

@@

Some taxonomic groups have microfossils that often come in groups of attached individuals. For example, for diatoms valves are counted, but these are commonly paired or, for some taxa, in long chains.
